---
title: "SHC 798 Assignment 1, 2025"
author: "Richard Lubega"
date: "`r Sys.Date()`"
output:
  pdf_document:
    latex_engine: xelatex
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# SHC 798 Assignment 1, 2025

## Part 3: Simple regression

### Question 3

##### (a)

```{r, Part-3-3-a}
time <- 0:9
count <- c(500, 400, 320, 250, 190, 150, 120, 90, 70, 50)
decay <- data.frame(time, count)

plot(decay$count, decay$time, type = "b", main = "Bacterial Counts at Different Times", pch=20, xlab = "Time (hours)", ylab = "Bacterial Count")

```

#### (b)

**Fit an exponential decay model and determine if this function better explains the data than a simple linear model**.

1.  **Simple Linear Model**

```{r, Part-3-3-b-1}
# Simple Linear Model
lm_decay <- lm(count ~ time, data = decay)
summary(lm_decay)
```

2.  **Exponential Model**

-   The exponential Model is obtained from a logged response model.
-   From general decay models, C(t) = C~0~ . e^−k . t^, where C implies bacterial count and t, time (hours).
-   We linearise to log[C(t)] = log[C~0~] − k . t, which is generally written as
    -   log(count) = β~0~ + β~1~ . time + E~*i*~ ............. a logged response model

```{r, Part-3-3-b-2}
exp_decay <- lm(log(count) ~ time, data = decay)
summary(exp_decay)
```

**Visualising the models**

```{r, Part-3-3-b-3}
plot(time, count, pch = 19, col = "black",
     xlab = "Time (hours)", ylab = "Bacterial Count",
     main = "Decay Model: Linear vs Exponential Fit")
abline(lm_decay, col = "red", lwd = 2)
# abline(exp_decay, col = "red", lwd = 2)

lines(time, exp(predict(exp_decay)), col = "blue", lwd = 2, lty = 2)
legend("topright", legend = c("Linear Fit", "Exponential Fit"),
       col = c("red", "blue"), lwd = 2, lty = c(1, 2))
```

**Comparing Explanatory Power**:

The exponential model *fits* (explains) the data **better** because it does have a higher R-Squared value (0.998) compared to the simple linear model (with 0.9299).

#### (c)

**Predict the bacterial count at time = 10 hours**

```{r, Part-3-3-c}
pred_c <- predict(exp_decay, newdata = data.frame(time = 10))
pred_10 <- exp(pred_c)
cat("the bacterial count at time = 10 hours is:", pred_10, "\n")
```

#### (d)

**Compute a 95% confidence interval for the estimated decay rate**

```{r, Part-3-3-d}
# The decay rate is the slope
cat("the 95% confidence interval for the estimated decay rate is: \n")
confint(exp_decay, "time")
```

